Multiproxy analysis exploring patterns of diet and disease in dental calculus and skeletal remains from a 19th century Dutch population

### 2024-04-24 analysis/paper/_methods.qmd
incoporate changes from pcj branch

<!-- methods for age, sex, and stature estimation -->
Demographic and pathological analyses were conducted in the Laboratory for Human
Osteoarchaeology at Leiden University. Sex was estimated using cranial
and pelvic morphological traits [@Standards1994]. Age-at-death was estimated using
dental wear,
auricular and pubic surface appearance, cranial suture closure, and epiphyseal fusion 
[@Standards1994; @SucheyBrooks1990; @buckberryAuricular2002; @lovejoyAuricular1985; @meindlSutureClosure1985],
and divided into the following categories:
early young adult (18-24 years), late young adult (25-34 years), middle adult (
35-49 years), old adult (50+ years). Preservation was visually scored on a
four-stage scale (excellent, good, fair, poor) based on the surface condition of
the bones and the extent of taphonomic degradation.

>>>>

<!-- methods for age, sex, and stature estimation -->
Demographic and pathological analyses were conducted in the Laboratory for Human
Osteoarchaeology at Leiden University. Sex was estimated using cranial
and pelvic morphological traits [@Standards1994]. Age-at-death was estimated using
dental wear,
auricular and pubic surface appearance, cranial suture closure, and epiphyseal fusion 
[@Standards1994; @SucheyBrooks1990; @buckberryAuricular2002; @lovejoyAuricular1985; @meindlSutureClosure1985],
and divided into the following categories:
early young adult (18-24 years), late young adult (25-34 years), middle adult
(35-49 years), old adult (50+ years). Preservation was visually scored on a
four-stage scale (excellent, good, fair, poor) based on the surface condition of
the bones and the extent of taphonomic degradation.

### 2024-04-24 analysis/paper/paper.qmd
incoporate changes from pcj branch

We wish to thank Kirsten Ziesemer for helping track down the calculus samples
from her studies. Additionally, we owe special thanks to Vincent Falger and Kees
de Groot from the Middenbeemster Historical Society for their input on early
draft manuscripts and a lovely guided tour of Middenbeemster. We also thank
Louise Le Meillour for taking on our preprint and providing helpful comments,
and Mario Zimmerman and two other anonymous reviewers for their time and invaluable
insights.

@@

All raw data is available on Zenodo (<https://doi.org/10.5281/zenodo.8061483>).
Analysis scripts, and the source code for the manuscript and
supplementary materials are available as a research compendium
(<https://doi.org/10.5281/zenodo.7649824>)
using the structure recommended by the **rrtools** R package [@Rrrtools].

>>>>

We wish to thank Kirsten Ziesemer for helping track down the calculus samples
from her studies. Additionally, we owe special thanks to Vincent Falger and Kees
de Groot from the Middenbeemster Historical Society for their input on early
draft manuscripts and a lovely guided tour of Middenbeemster.

We also thank Louise Le Meillour for taking on our preprint and providing helpful
comments, and Mario Zimmerman and two other anonymous reviewers for their time
and invaluable insights. This paper was evaluated during a CODECHECK event at TU
Delft by Daniel Nüst and Stephen Eglen, who provided helpful feedback on the
reproducibility of all code used in this paper.

Preprint version 5 of this article has been peer-reviewed and recommended by PCI
Archaeology <https://doi.org/10.24072/pci.archaeo.100389> [@meillourDetectionPlantderived2023].

@@

All raw data is available on Zenodo [<https://doi.org/10.5281/zenodo.8061483>,
@bartholdyDentalPathological2023].
Analysis scripts, and the source code for the manuscript and
supplementary materials are available as a research compendium
[<https://doi.org/10.5281/zenodo.7649824>, @bartholdyMB11CalculusPilot]
using the structure recommended by the **rrtools** R package [@Rrrtools].
